---
title: "Fitting Multiple taxa with Generalized additive models (gam)"
author: "Nengjun Yi, Xinyan Zhang"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{mgam}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
---

```{r, SETTINGS-knitr, include=FALSE}
stopifnot(require(knitr))
opts_chunk$set(
  comment=NA, 
  message = FALSE, 
  warning = FALSE, 
  eval = params$EVAL,
  dev = "png",
  dpi = 500,
  fig.asp = 0.618,
  fig.width = 14,
  out.width = "100%",
  fig.align = "center"
)
```

```{r, pkgs, include=FALSE}
library("NBZIMM")
```


## Introduction

This vignette focuses on demonstrations of fitting generalized additive models for multiple longitudinal microbiome taxa with **NBZIMM**. 

The **NBZIMM**  provides a `mgam` functions for setting up and fitting generalized additive models by working with the function `gam` from R package **mgcv**. 

In this vignette we'll use the `mgam` function in the 
**NBZIMM** package to demonstrate the analysis of a public available dataset from Romero et al (2014).

## Demonstration with Example

This example used a published dataset from [Romero et al.(2014)](https://microbiomejournal.biomedcentral.com/articles/10.1186/2049-2618-2-4). It was a case-control longitudinal study collected 22 pregnant women who delivered at term and 32 non-pregnant women, each measured at multiple time points. The data consist of two data components: OTU and SampleData. OTU contains microbiome count data for 900 samples with 143 taxa; SampleData contains data of sample variables for all the samples with 9 variables, including subject ID, pregnant status, total sequencing read, age, race, etc.

Here we import the dataset from the package **NBZIMM** and check the clinical variables first.


```{r, romero}
data(Romero) # see help("Romero")
names(Romero)
```

Get the necessary variables to fit the models
```{r, variables}
otu = Romero$OTU; dim(otu)
sam = Romero$SampleData; dim(sam)
colnames(sam)

N = sam[, "Total.Read.Counts"]        
Days = sam$GA_Days
Age = sam$Age
Race = sam$Race
preg = sam$pregnant; table(preg)
subject = sam[, "Subect_ID"]; table(subject)

```

### analyze all taxa using generalized additive models with a given nonzero proportion

```{r, mgam}
f = mgam(y=otu, formula= ~ offset(log(N)) + Days + Age + Race + preg + s(subject, bs="re"), 
         family=nb(), min.p=0.2)
out = fixed(f)

para = out$parametric_terms
para = para[para[,2]!="(Intercept)", ]
par(mfrow = c(1, 1), cex.axis = 1, mar = c(2, 12, 4, 4))
plot.fixed(res=para[,c("Estimate","Std.Error","padj")], 
           threshold=0.001, gap=500, col.pts=c("black", "grey"),
           cex.axis=0.6, cex.var=0.7)

preg.coefs = para[para$variables=="preg",]
par(mfrow = c(1, 1), cex.axis = 1, mar = c(2, 10, 4, 4))
plot.fixed(res=preg.coefs[,c("Estimate","Std.Error","padj")], 
           threshold=0.05, gap=300, main="Covariate: pregnant",
           cex.axis=0.6, cex.var=0.7)

g = heat.p(df=para, p.breaks = c(0.001, 0.01, 0.05), 
       colors = c("black", "darkgrey", "grey", "lightgrey"),
       zigzag=c(T,F), abbrv=c(T,F), margin=c(1,6), y.size=8,
       legend=T)
```



